---
title: "Analysis of Recount data"
output: html_document
author: Jeff Leek
---

`r library(knitr); opts_chunk$set(cache=TRUE)`

### Load packages

You will need the RSkittleBrewer package for this vignette to run. Installation instructions are available here:

* https://github.com/alyssafrazee/RskittleBrewer

This code also accesses data from the ReCount database:

* http://bowtie-bio.sourceforge.net/recount/

You will also need R version 3.1.0 or greater and Bioconductor 3.0 or greater. The zebrafishRNASeq package might need to be installed from source. These analyses are based on the devel version of sva (version 3.11.2 or greater).

```{r load,message=FALSE}
library(zebrafishRNASeq)
library(RSkittleBrewer)
library(genefilter)
library(RUVSeq)
library(edgeR)
library(sva)
library(ffpe)
library(RColorBrewer)
library(corrplot)
library(limma)
trop = RSkittleBrewer('tropical')
```

## Read data from ReCount

```{r recount, dependson="load"}
load(url("http://bowtie-bio.sourceforge.net/recount/ExpressionSets/montpick_eset.RData"))
mpdat= exprs(montpick.eset)
mppheno = pData(montpick.eset)
mpdat = mpdat[rowMeans(mpdat) > 5,]
```

## Read pedigree info from Hapmap
```{r pedinfo, dependson="recount"}

# CEU

z = gzcon(url("http://hapmap.ncbi.nlm.nih.gov/downloads/samples_individuals/pedinfo2sample_CEU.txt.gz"))
raw = textConnection(readLines(z))
close(z)
ceuped = read.table(raw)
close(raw)


## YRI

z = gzcon(url("http://hapmap.ncbi.nlm.nih.gov/downloads/samples_individuals/pedinfo2sample_YRI.txt.gz"))
raw = textConnection(readLines(z))
close(z)
yriped = read.table(raw)
close(raw)

```

## Match sex info with samples

Remove all samples where we can not identify sex information exactly.
```{r match, dependson="pedinfo"}
mppheno$sex = rep(NA,129)
for(i in 1:dim(mppheno)[1]){
  if(mppheno$population[i] == "CEU"){
    ind = grep(mppheno$sample.id[i],ceuped[,7])
    if(length(ind)==1){
      mppheno$sex[i] = ceuped[ind,5]
    }
  }
  if(mppheno$population[i] == "YRI"){
    ind = grep(mppheno$sample.id[i],yriped[,7])
    if(length(ind)==1){
      mppheno$sex[i] = yriped[ind,5]
    }
  }  
}

dat0 = mpdat[,!is.na(mppheno$sex)]
mppheno = mppheno[!is.na(mppheno$sex),]
group = as.numeric(mppheno$sex)
study = as.numeric(mppheno$study)
table(study,group)
```

## Estimate latent factors with different methods

```{r estbatch,dependson="match"}
## Set null and alternative models (ignore batch)
mod1 = model.matrix(~group)
mod0 = cbind(mod1[,1])

## Estimate batch with svaseq (unsupervised)
batch_unsup_sva = svaseq(dat0,mod1,mod0,n.sv=1)$sv

## Estimate batch with pca
ldat0 = log(dat0 + 1)
batch_pca = svd(ldat0 - rowMeans(ldat0))$v[,1]


## Estimate batch with ruv (residuals)
## this procedure follows the RUVSeq vignette
## http://www.bioconductor.org/packages/devel/bioc/vignettes/RUVSeq/inst/doc/RUVSeq.pdf

x <- as.factor(group)
design <- model.matrix(~x)
y <- DGEList(counts=dat0, group=x)
y <- calcNormFactors(y, method="upperquartile")
y <- estimateGLMCommonDisp(y, design)
y <- estimateGLMTagwiseDisp(y, design)
fit <- glmFit(y, design)
res <- residuals(fit, type="deviance")
seqUQ <- betweenLaneNormalization(dat0, which="upper")
controls = rep(TRUE,dim(dat0)[1])
batch_ruv_res = RUVr(seqUQ,controls,k=1,res)$W

## Estimate batch with ruv empirical controls
## this procedure follows the RUVSeq vignette
## http://www.bioconductor.org/packages/devel/bioc/vignettes/RUVSeq/inst/doc/RUVSeq.pdf

y <- DGEList(counts=dat0, group=x)
y <- calcNormFactors(y, method="upperquartile")
y <- estimateGLMCommonDisp(y, design)
y <- estimateGLMTagwiseDisp(y, design)

fit <- glmFit(y, design)
lrt <- glmLRT(fit, coef=2)

controls = rank(lrt$table$LR) <= 400
batch_ruv_emp <- RUVg(dat0, controls, k=1)$W
```

### Plot batch estimates

```{r plotbatch,dependson="estbatch"}
## Plot the results
plot(batch_unsup_sva ~ mppheno$study,pch=19,col=trop[1],main="unsupervised sva")
plot(batch_pca ~ mppheno$study,pch=19,col=trop[2],main="pca")
plot(batch_ruv_res ~ mppheno$study,pch=19,col=trop[3],main="residual ruv")
plot(batch_ruv_emp ~ mppheno$study,pch=19,col=trop[3],main="empirical controls ruv")
```


## Plot absolute correlation between batch estimates

```{r corbatch,dependson="estbatch"}
batchEstimates = cbind(group,study,batch_unsup_sva,batch_pca,
                       batch_ruv_res,batch_ruv_emp)
colnames(batchEstimates) = c("group","study","usva","pca","ruvres","ruvemp")

corr = abs(cor(batchEstimates))
corr
cols = colorRampPalette(c(trop[2],"white",trop[1]))
par(mar=c(5,5,5,5))
corrplot(corr,method="ellipse",type="lower",col=cols(100),tl.pos="d")
```

## Compare results with different methods to just using study

```{r comparede, dependson="estbatch",fig.align="center",fig.height=7,fig.width=7}
dge <- DGEList(counts=dat0)
dge <- calcNormFactors(dge)
catplots = tstats = vector("list",6)
adj = c("+ study", "+ batch_unsup_sva",
         "+ batch_ruv_res", "+ batch_ruv_emp",
        "+ batch_pca","")


for(i in 1:6){
  design = model.matrix(as.formula(paste0("~ group",adj[i])))
  v <- voom(dge,design,plot=FALSE)
  fit <- lmFit(v,design)
  fit <- eBayes(fit)
  tstats[[i]] = abs(fit$t[,2])
  names(tstats[[i]]) = as.character(1:dim(dat0)[1])
  catplots[[i]] = CATplot(-rank(tstats[[i]]),-rank(tstats[[1]]),maxrank=1000,make.plot=F)
}
```

## Make the CATplot

```{r catplot, dependson="comparede"}

plot(catplots[[2]],ylim=c(0,1),col=trop[2],lwd=3,type="l",ylab="Concordance Between using study and different methods",xlab="Rank")
lines(catplots[[3]],col=trop[1],lwd=3,lty=2)
lines(catplots[[4]],col=trop[1],lwd=3)
lines(catplots[[5]],col=trop[3],lwd=3,lty=3)
lines(catplots[[6]],col=trop[4],lwd=3)

legend(200,0.5,legend=c("Unsup. svaseq","RUV Res.", "RUV Emp.","PCA","No adjustment"),col=trop[c(2,1,1,3,4)],lty=c(1,2,1,3,1),lwd=3)
```


# Now try an unbalanced design

```{r unbal,dependson="match"}
set.seed(35353)
index = c(sample(which(group==1 & study==1),size=20),sample(which(group==1 & study==2),size=10),
          sample(which(group==2 & study==1),size=10),sample(which(group==2 & study==2),size=20))
dat0 = dat0[,index]
group = group[index]
study = study[index]
```


## Estimate latent factors with different methods

```{r estbatch2,dependson="unbal"}
## Set null and alternative models (ignore batch)
mod1 = model.matrix(~group)
mod0 = cbind(mod1[,1])

## Estimate batch with svaseq (unsupervised)
batch_unsup_sva = svaseq(dat0,mod1,mod0,n.sv=1)$sv

## Estimate batch with pca
ldat0 = log(dat0 + 1)
batch_pca = svd(ldat0 - rowMeans(ldat0))$v[,1]


## Estimate batch with ruv (residuals)
## this procedure follows the RUVSeq vignette
## http://www.bioconductor.org/packages/devel/bioc/vignettes/RUVSeq/inst/doc/RUVSeq.pdf

x <- as.factor(group)
design <- model.matrix(~x)
y <- DGEList(counts=dat0, group=x)
y <- calcNormFactors(y, method="upperquartile")
y <- estimateGLMCommonDisp(y, design)
y <- estimateGLMTagwiseDisp(y, design)
fit <- glmFit(y, design)
res <- residuals(fit, type="deviance")
seqUQ <- betweenLaneNormalization(dat0, which="upper")
controls = rep(TRUE,dim(dat0)[1])
batch_ruv_res = RUVr(seqUQ,controls,k=1,res)$W

## Estimate batch with ruv empirical controls
## this procedure follows the RUVSeq vignette
## http://www.bioconductor.org/packages/devel/bioc/vignettes/RUVSeq/inst/doc/RUVSeq.pdf

y <- DGEList(counts=dat0, group=x)
y <- calcNormFactors(y, method="upperquartile")
y <- estimateGLMCommonDisp(y, design)
y <- estimateGLMTagwiseDisp(y, design)

fit <- glmFit(y, design)
lrt <- glmLRT(fit, coef=2)

controls = rank(lrt$table$LR) <= 400
batch_ruv_emp <- RUVg(dat0, controls, k=1)$W
```

### Plot batch estimates

```{r plotbatch2,dependson="estbatch2"}
## Plot the results
plot(batch_unsup_sva ~ mppheno$study[index],pch=19,col=trop[1],main="unsupervised sva")
plot(batch_pca ~ mppheno$study[index],pch=19,col=trop[2],main="pca")
plot(batch_ruv_res ~ mppheno$study[index],pch=19,col=trop[3],main="residual ruv")
plot(batch_ruv_emp ~ mppheno$study[index],pch=19,col=trop[3],main="empirical controls ruv")
```


## Plot absolute correlation between batch estimates

```{r corbatch2,dependson="estbatch2"}
batchEstimates = cbind(group,study,batch_unsup_sva,batch_pca,
                       batch_ruv_res,batch_ruv_emp)
colnames(batchEstimates) = c("group","study","usva","pca","ruvres","ruvemp")

corr = abs(cor(batchEstimates))
corr
cols = colorRampPalette(c(trop[2],"white",trop[1]))
par(mar=c(5,5,5,5))
corrplot(corr,method="ellipse",type="lower",col=cols(100),tl.pos="d")
```

## Compare results with different methods to just using study

```{r comparede2, dependson="estbatch2",fig.align="center",fig.height=7,fig.width=7}
dge <- DGEList(counts=dat0)
dge <- calcNormFactors(dge)
catplots = tstats = vector("list",6)
adj = c("+ study", "+ batch_unsup_sva",
         "+ batch_ruv_res", "+ batch_ruv_emp",
        "+ batch_pca","")


for(i in 1:6){
  design = model.matrix(as.formula(paste0("~ group",adj[i])))
  v <- voom(dge,design,plot=FALSE)
  fit <- lmFit(v,design)
  fit <- eBayes(fit)
  tstats[[i]] = abs(fit$t[,2])
  names(tstats[[i]]) = as.character(1:dim(dat0)[1])
  catplots[[i]] = CATplot(-rank(tstats[[i]]),-rank(tstats[[1]]),maxrank=1000,make.plot=F)
}
```

## Make the CATplot

```{r catplot2, dependson="comparede2"}

plot(catplots[[2]],ylim=c(0,1),col=trop[2],lwd=3,type="l",ylab="Concordance Between using study and different methods",xlab="Rank")
lines(catplots[[3]],col=trop[1],lwd=3,lty=2)
lines(catplots[[4]],col=trop[1],lwd=3)
lines(catplots[[5]],col=trop[3],lwd=3,lty=3)
lines(catplots[[6]],col=trop[4],lwd=3)



legend(200,0.5,legend=c("Unsup. svaseq","RUV Res.", "RUV Emp.","PCA","No adjustment"),col=trop[c(2,1,1,3,4)],lty=c(1,2,1,3,1),lwd=3)
```




### Session Info

```{r}
sessionInfo()
```


