---
title: "Analysis of Zebrafish data"
output: html_document
author: Jeff Leek
---

`r library(knitr); opts_chunk$set(cache=TRUE)`

### Load packages

You will need the RSkittleBrewer package for this vignette to run. Installation instructions are available here:

* https://github.com/alyssafrazee/RskittleBrewer

You will also need R version 3.1.0 or greater and Bioconductor 3.0 or greater. The zebrafishRNASeq package might need to be installed from source. These analyses are based on the devel version of sva (version 3.11.2 or greater).

```{r load,message=FALSE}
library(zebrafishRNASeq)
library(RSkittleBrewer)
library(genefilter)
library(RUVSeq)
library(edgeR)
library(sva)
library(ffpe)
library(RColorBrewer)
library(corrplot)
library(limma)
trop = RSkittleBrewer('tropical')
```

### Load and process the zebrafish data

The first comparison will be on the zebrafish data like those in the RUVSeq vignette. For the vignette code see this link http://bioconductor.org/packages/devel/bioc/vignettes/RUVSeq/inst/doc/RUVSeq.R.

```{r zfdata,dependson="loadLibraries"}
data(zfGenes)
filter = apply(zfGenes, 1, function(x) length(x[x>5])>=2)
filtered = zfGenes[filter,]
genes = rownames(filtered)[grep("^ENS", rownames(filtered))]
controls = grepl("^ERCC", rownames(filtered))
spikes =  rownames(filtered)[grep("^ERCC", rownames(filtered))]
group = as.factor(rep(c("Ctl", "Trt"), each=3))
set = newSeqExpressionSet(as.matrix(filtered),
                           phenoData = data.frame(group, row.names=colnames(filtered)))
dat0 = counts(set)
```

## Estimate latent factors with different methods

```{r estbatch,dependson="zfdata"}
## Set null and alternative models (ignore batch)
mod1 = model.matrix(~group)
mod0 = cbind(mod1[,1])

## Estimate batch with svaseq (unsupervised)
batch_unsup_sva = svaseq(dat0,mod1,mod0,n.sv=1)$sv

## Estimate batch with svaseq (supervised)
batch_sup_sva = svaseq(dat0,mod1,mod0,controls=controls,n.sv=1)$sv

## Estimate batch with pca
ldat0 = log(dat0 + 1)
batch_pca = svd(ldat0 - rowMeans(ldat0))$v[,1]

## Estimate batch with ruv (controls known)
batch_ruv_cp <- RUVg(dat0, cIdx= spikes, k=1)$W

## Estimate batch with ruv (residuals)
## this procedure follows the RUVSeq vignette
## http://www.bioconductor.org/packages/devel/bioc/vignettes/RUVSeq/inst/doc/RUVSeq.pdf

x <- as.factor(group)
design <- model.matrix(~x)
y <- DGEList(counts=dat0, group=x)
y <- calcNormFactors(y, method="upperquartile")
y <- estimateGLMCommonDisp(y, design)
y <- estimateGLMTagwiseDisp(y, design)
fit <- glmFit(y, design)
res <- residuals(fit, type="deviance")
seqUQ <- betweenLaneNormalization(dat0, which="upper")
controls = rep(TRUE,dim(dat0)[1])
batch_ruv_res = RUVr(seqUQ,controls,k=1,res)$W

## Estimate batch with ruv empirical controls
## this procedure follows the RUVSeq vignette
## http://www.bioconductor.org/packages/devel/bioc/vignettes/RUVSeq/inst/doc/RUVSeq.pdf

y <- DGEList(counts=dat0, group=x)
y <- calcNormFactors(y, method="upperquartile")
y <- estimateGLMCommonDisp(y, design)
y <- estimateGLMTagwiseDisp(y, design)

fit <- glmFit(y, design)
lrt <- glmLRT(fit, coef=2)

controls=rank(lrt$table$LR) <= 400
batch_ruv_emp <- RUVg(dat0, controls, k=1)$W
```

### Plot batch estimates

```{r plotbatch,dependson="estbatch"}
## Plot the results
plot(batch_unsup_sva,pch=19,col=trop[1],main="unsupervised sva")
plot(batch_sup_sva,pch=19,col=trop[1],main="supervised sva")
plot(batch_pca,pch=19,col=trop[2],main="pca")
plot(batch_ruv_cp,pch=19,col=trop[3],main="control probes ruv")
plot(batch_ruv_res,pch=19,col=trop[3],main="residual ruv")
plot(batch_ruv_emp,pch=19,col=trop[3],main="empirical controls ruv")
```


## Plot absolute correlation between batch estimates

```{r corbatch,dependson="estbatch"}

batchEstimates = cbind(group,batch_unsup_sva,batch_sup_sva,
                       batch_pca,batch_ruv_cp,batch_ruv_res,batch_ruv_emp)
colnames(batchEstimates) = c("group","usva","ssva","pca","ruvcp","ruvres","ruvemp")

corr = abs(cor(batchEstimates))
cols = colorRampPalette(c(trop[2],"white",trop[1]))
par(mar=c(5,5,5,5))
corrplot(corr,method="ellipse",type="lower",col=cols(100),tl.pos="d")
```

## Compare results with different methods to supervised svaseq

```{r comparede, dependson="estbatch",fig.align="center",fig.height=7,fig.width=7}
dge <- DGEList(counts=dat0)
dge <- calcNormFactors(dge)
catplots = tstats = vector("list",7)
adj = c( "+ batch_sup_sva","+ batch_unsup_sva",
         "+ batch_ruv_cp", "+ batch_ruv_res",
         "+ batch_ruv_emp","+ batch_pca","")


for(i in 1:7){
  design = model.matrix(as.formula(paste0("~ group",adj[i])))
  v <- voom(dge,design,plot=FALSE)
  fit <- lmFit(v,design)
  fit <- eBayes(fit)
  tstats[[i]] = abs(fit$t[,2])
  names(tstats[[i]]) = as.character(1:dim(dat0)[1])
  catplots[[i]] = CATplot(-rank(tstats[[i]]),-rank(tstats[[1]]),maxrank=400,make.plot=F)
}

plot(catplots[[2]],ylim=c(0,1),col=trop[1],lwd=3,type="l",ylab="Concordance Between Supervised svaseq and other methods",xlab="Rank")
lines(catplots[[3]],col=trop[2],lwd=3)
lines(catplots[[4]],col=trop[2],lwd=3,lty=2)
lines(catplots[[5]],col=trop[2],lwd=3,lty=3)
lines(catplots[[6]],col=trop[3],lwd=3)
lines(catplots[[7]],col=trop[4],lwd=3)


legend(200,0.5,legend=c("Unsup. svaseq","RUV CP","RUV Res.", "RUV Emp.","PCA", "No adjustment"),col=trop[c(1,2,2,2,3,4)],lty=c(1,1,2,3,1,1),lwd=3)
```



## Compare results with different methods to unsupervised svaseq

```{r comparede2, dependson="estbatch",fig.align="center",fig.height=7,fig.width=7}
dge <- DGEList(counts=dat0)
dge <- calcNormFactors(dge)
catplots = tstats = vector("list",7)
adj = c( "+ batch_unsup_sva","+ batch_sup_sva",
         "+ batch_ruv_cp", "+ batch_ruv_res",
         "+ batch_ruv_emp","+ batch_pca","")


for(i in 1:7){
  design = model.matrix(as.formula(paste0("~ group",adj[i])))
  v <- voom(dge,design,plot=FALSE)
  fit <- lmFit(v,design)
  fit <- eBayes(fit)
  tstats[[i]] = abs(fit$t[,2])
  names(tstats[[i]]) = as.character(1:dim(dat0)[1])
  catplots[[i]] = CATplot(-rank(tstats[[i]]),-rank(tstats[[1]]),maxrank=400,make.plot=F)
}

plot(catplots[[2]],ylim=c(0,1),col=trop[1],lwd=3,type="l",ylab="Concordance Between Unsupervised svaseq and other methods",xlab="Rank")
lines(catplots[[3]],col=trop[2],lwd=3,lty=2)
lines(catplots[[4]],col=trop[2],lwd=3,lty=1)
lines(catplots[[5]],col=trop[2],lwd=3,lty=3)
lines(catplots[[6]],col=trop[3],lwd=3)
lines(catplots[[7]],col=trop[4],lwd=3)


legend(200,0.5,legend=c("Sup. svaseq","RUV CP","RUV Res.", "RUV Emp.","PCA", "No adjustment"),col=trop[c(1,2,2,2,3,4)],lty=c(1,2,1,3,1,1),lwd=3)
```

### Session Info

```{r}
sessionInfo()
```




